\documentclass[10pt]{amsart}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[multiple]{footmisc}
\usepackage{multicol}
%\usepackage{enumerate}
%\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{mathdots}

%let's have nice pdf metadata :)
\usepackage[pdfauthor={Kameryn J Williams},
    pdftitle={Math335 Chapter 1 Problem Set},
    hidelinks 
]{hyperref}


\input{../../macros.tex}


%\theoremstyle{theorem}
\ifdefined\project \else \newtheorem{project}[theorem]{Project Idea} \fi
\ifdefined\theoremschema \else \newtheorem{theoremschema}[theorem]{Theorem Schema} \fi



\title{Math 355 problem set \\ Chapter 1: Ordinals}

\author{Kameryn J. Williams}
\address[Kameryn J. Williams]{
Bard College at Simon's Rock \\
84 Alford Rd \\
Great Barrington, MA 01230}
\email{kwilliams@simons-rock.edu}
\urladdr{http://kamerynjw.net}

\date{\today}

\begin{document}

\maketitle

%% always do 10 problems per set

\begin{problem}
Let $A$ be a set. Consider the collection of all ordinal numbers of well-orders whose domain is a subset of $A$, ordered by $\le$. Call this collection of ordinals $\aleph(A)$. Prove that there is no injection $\aleph(A) \to A$. As a corollary derive that there is a well-order whose domain is not countable.
\end{problem}

\begin{problem}
Prove that a linear order satisfying induction implies it is well-founded. That is, consider a linear order $(X,\le)$. Assume that for any $Y \subseteq X$ if (for all  $x \in X$ we have [if every $y < x$ is in $X$ then $x \in X$]), then $Y = X$. (Brackets added to make the logical structure clear.) Prove that $X$ must be well-founded.
\end{problem}

\begin{problem}
Prove that the definition of ordinal addition by transfinite recursion matches our first definition of ordinal addition.
\end{problem}

\begin{problem}
Prove that the definition of ordinal multiplication by transfinite recursion matches our first definition of ordinal multiplication.
\end{problem}


%% prove some properties of ordinal sum mult and exp
\begin{problem}
Use the recursive definition of ordinal arithmetic to prove these properties.
\begin{align*}
0 \cdot \alpha &= 0; \\
1 \cdot \alpha &= \alpha; \\
(\alpha \cdot \beta) \cdot \gamma &= \alpha \cdot (\beta \cdot \gamma); \\
\alpha \cdot (\beta + \gamma) &= \alpha \cdot \beta + \alpha \cdot \gamma;
\end{align*}
\end{problem}

%% some more properties of ordinal arithmetic
\begin{problem}
Prove these properties of ordinal exponentation.
\begin{align*}
0^\beta &= 0, \qquad\qquad\text{if } \beta \ne 0; \\
1^\beta &= 1; \\
\alpha^\beta &< \alpha^\gamma, \qquad\qquad\text{if } 1 < \alpha \mand \beta < \gamma; \\
\alpha^\gamma &\le \beta^\gamma, \qquad\qquad\text{if } \alpha \le \beta.
\end{align*}
\end{problem}



\begin{problem}
Prove these properties of ordinal exponentiation:
\begin{align*}
\alpha^{\beta + \gamma} &= \alpha^\beta \cdot \alpha^\gamma; \\
\alpha^{\beta \cdot \gamma} &= \left(\alpha^\beta\right)^\gamma.
\end{align*}
\end{problem}

\begin{problem}
Prove that every ordinal can be written uniquely in the form
\[
\omega^{\alpha_0} \cdot n_0 + \omega^{\alpha_1} \cdot n_1 + \cdots + \omega^{\alpha_k} \cdot n_k,
\]
where $k$ is finite, each $n_0$ is a finite ordinal, and $\alpha_0 > \alpha_1 > \cdots > \alpha_k$ are ordinals. 
\end{problem}


\begin{problem}
A function $f$ on the ordinals is \emph{increasing} if $\alpha \le f(\alpha)$ and $f$ is \emph{continuous} if given any sequence $\seq{\alpha_i : i \in \lambda}$ we have
\[
f\left(\sup_{i \in \lambda} \alpha_i\right) = \sup_{i \in \lambda} f(\alpha_i).
\]
Show that if $f : \Ord \to \Ord$ is increasing and continuous then there always is an ordinal $\alpha$ so that $f(\alpha) = \alpha$. Moreover show that if $\beta$ is any ordinal we can always find such $\alpha$ to be larger than $\beta$.
\end{problem}


\begin{problem}
\emph{[This problem uses some ideas from real analysis or topology. If you've taken neither of those classes you may find it challenging.]}

A set $C \subseteq \Rbb$ is called \emph{closed} if it contais all its accumulation points, where $x$ is an \emph{accumulation point} of $C$ if there are points in $C \setminus \{x\}$ abitrarily close to $x$.\footnote{That is, for any distance $\epsilon > 0$ there is $y \in C \setminus \{x\}$ so that $\abs{x - y} < \epsilon$. If you know another characterization of closed sets you're welcome to use it instead for this problem.}
A set $P \subseteq \Rbb$ is called \emph{perfect} if it is equal to the set of its accumulation points. Follow this outline to prove that any closed set $C \subseteq \Rbb$ can be decomposed into a disjoint union of a perfect set and a countable set.
\begin{enumerate}
\item Define the \emph{derivative} of a set $A \subseteq \Rbb$ to be $A' = \{ x \in \Rbb : x$ is an accumulation point of $A\}$. Show that if $A$ is closed then so is $A'$.
\item Starting from a closed set $C$ recursively define a sequence of closed sets $C_\alpha$:
  \begin{itemize}
  \item $C_0 = C$;
  \item $C_{\alpha+1} = C_\alpha'$; and
  \item If $\gamma$ is limit, then $C_\gamma = \bigcup_{\alpha < \gamma} C_\alpha$.
  \end{itemize}
Say why every $C_\alpha$ is closed and that if $C_\alpha = C_{\alpha+1}$ then $C_\alpha$ is perfect.
\item Argue that $C_\alpha \setminus C_{\alpha+1}$ is always countable. [Hint: First see that if $x \in X \setminus X'$ then there is an open interval $U \ni x$ so that $U \cap X = \{x\}$.]
\item Finally, show that no matter what closed $C$ you start with, there's a countable ordinal $\alpha$ so that $C_\alpha = C_{\alpha+1}$.
\item Conclude that $C$ is the union of a perfect set and a countable set.
\end{enumerate}
\end{problem}


\end{document}

