\documentclass[10pt]{amsart}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[multiple]{footmisc}
\usepackage{multicol}
%\usepackage{enumerate}
%\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{mathdots}

%let's have nice pdf metadata :)
\usepackage[pdfauthor={Kameryn J Williams},
    pdftitle={Math335 Chapter 3 Problem Set},
    hidelinks 
]{hyperref}


\input{../../macros.tex}


%\theoremstyle{theorem}
\ifdefined\project \else \newtheorem{project}[theorem]{Project Idea} \fi
\ifdefined\theoremschema \else \newtheorem{theoremschema}[theorem]{Theorem Schema} \fi



\title{Math 355 problem set \\ Chapter 3: The cumulative hierarchy}

\author{Kameryn J. Williams}
\address[Kameryn J. Williams]{
Bard College at Simon's Rock \\
84 Alford Rd \\
Great Barrington, MA 01230}
\email{kwilliams@simons-rock.edu}
\urladdr{http://kamerynjw.net}

\date{\today}

\begin{document}

\maketitle

%% always do 10 problems per set

%% 2 more

Let $t$ be a transitive set. When we say that an axiom is \emph{true in $t$} we mean that if you restrict the quantifiers in the axiom to only quantify over elements of $t$ then it comes out as true. The idea is, you imagine that $t$ were the whole universe you're quantifying over.

\begin{problem}
Check that the axioms of Extensionality, Union, and Infinity are true in $\omega_1$, but the axioms of Pairing and Powerset are not true in $\omega_1$.
\end{problem}

\begin{problem}
Suppose $x \in \Vrm_\alpha$. Prove that any choice function on $x$ is in $\Vrm_{\alpha+n}$ for some finite $n$. What is the optimal value of $n$ you can achieve? Conclude that if $\gamma$ is limit then $\AC$ is true in $\Vrm_\gamma$.
\end{problem}

\begin{problem}
Prove that if $\gamma > \omega$ is limit then Foundation plus every axiom of $\Zsf$ is true in $\Vrm_\gamma$. 
\end{problem}

The last two problems together imply that the only sticking point to $\Vrm_\gamma$ satisfying all of $\ZFC$ is Replacement.

\begin{problem}
Show that Replacement is not true in $\Vrm_{\omega_1}$ nor in $\Vrm_{\aleph_\omega}$. [Hint: For each index $\alpha$ define a function from a set $\in \Vrm_{\alpha}$ which is onto $\alpha$. Explain why its image can't be a set in $\Vrm_\alpha$.]
\end{problem}

The next two problems ask you to prove the full Mostowski collapse theorem.

\begin{problem}
Prove the following: Suppose $(X,E)$ is a well-founded relation on a set $X$. Then there is a transitive set $t$ and an onto map $\pi : X \to t$ so that $x \mathbin E y$ if and only if $\pi(x) \in \pi(y)$.
\end{problem}

\begin{problem}
{[This problem builds on the previous one.]}
Say that a relation $(X,E)$ is \emph{extensional} if it satisfies the axiom of Extensionality. That is, $x,y \in X$ are equal if and only if for all $z \in X$ we have $z \mathbin E x$ iff $z \mathbin E y$. Prove that if $(X,E)$ is extensional and well-founded then the $\pi$ from the previous problem is a bijection.
\end{problem}

\begin{problem}
Prove that Replacement is equivalent to transfinite recursion. One direction we did in class, so what you need to prove is: working over the axioms $\Zsf + \AC$, prove that if transfinite recursion is valid then the image of a set under a class function is a set. [Hint: construct the image by transfinite recursion, using $\AC$ to have a well-order to work over.]
\end{problem}

The next two problems ask you to prove that Zorn's lemma is equivalent to the axiom of Choice.

\begin{problem}
Prove over $\ZFC$ that
\begin{itemize}
\item (Zorn) Suppose $(P,\le)$ is a partially ordered set with the property that if $C \subseteq P$ is linearly ordered by $\le$ then there is $b \in P$ so that $c \le b$ for all $c \in C$. Then there is a maximal element in $P$. Namely, there is $m \in P$ so that there is no $x \in P$ with $m < x$.
\end{itemize}
\end{problem}

\begin{problem}
Prove over $\ZF$ that (Zorn) implies $\AC$. 
\end{problem}


\begin{problem}
Prove that, over $\ZF$, $\AC$ is equivalent to
\begin{itemize}
\item There are no sets $A,B$ so that if $\kappa < \card A$ then $\kappa < \card B$ but $A$ and $B$ are incomparable in cardinality. (Meaning that there is no injection $A \to B$ and also no injection $B \to A$.
\end{itemize}
[Hint: This problem is a sequel to problem 6 from Chapter 2, which in turn was a sequel to problem 1 from Chapter 1. Do them first, or at least understand what they are saying.]
\end{problem}




\end{document}

